---
title: "Lecture 11 - ANOVA: Simple, nested, and factorial"
author: "Matthew E. Aiello-Lammens"
output: 
  html_document: 
    toc: yes
---

# ANOVA Review

**See written notes.**

#### Challenge

This is a review from last week. Construct and ANOVA model to test the effects of zinc concentration on diatom diversity in Rocky Mountain creeks.

Recall: From Medley and Clements (1998), by way of Logan 2010 - investigation of the impact of zinc contamination on diversity of diatom species in the USA Rocky Mountains.

```{r}
diatoms <- read.csv(file = "https://mlammens.github.io/Biostats/data/Logan_Examples/Chapter10/Data/medley.csv")
```


```{r}
## Look at the data
head(diatoms)

summary(aov(data = diatoms, DIVERSITY ~ ZINC ))

## Make the linear model
diatom_lm <- lm(data = diatoms, DIVERSITY ~ ZINC )

## Look at the ANOVA results
anova(diatom_lm)

## Look at the linear model results
summary(diatom_lm)

## Look at the model diagnostics
plot(diatom_lm)

```

## Post-hoc comparisons

Now we can say that the biodiversity measures taken at different zinc concentrations **DO NOT** come from a common distribution, but we don't know which are different from each other.

We could compare each concentration to each of the others separately, but recall the problems with multiple comparions. 
Instead, we'll use a *post-hoc* comparison method. 
Specifically, we'll use the Tukey Honest Significant Differences (HSD), but there are several other methods you might use as well (see Logan p. 265). 

```{r}
TukeyHSD(aov(data = diatoms, DIVERSITY ~ ZINC ))
```

Note, that for reasons that have never been clear to me, you can only use the TukeyHSD method when using the `aov` function.
For an alternative approach, see the `glht` function in the `multcomp` package. 
This is described in Logan (e.g., p. 267).

## Planned comparisons

Where as unplanned, post-hoc comparisons require a correction for multiple comparisons, planned comparisons do not. 
We will not cover planned comparisons in class, but they may show up in the homework or on the final. 
Example 10B in Logan provides an example of the use of planned comparisons.

## Non-parametric ANOVA

As with many of the other methods we've discussed, there are also non-parameteric ANOVA approachs. 
The most common is the Kruskal-Wallace test, which uses **rank based** statistics.


# Nested ANOVA

Sometimes your replicates aren't quite independent.

**See visual as per Logan, Fig 11.1.**

For formula, see Logan p. 284 and for calculation of F-ratios see Table 11.1.
**Note that Mean-Squared Errors are the same if B is random or fixed, but F-ratio is differen! Why? Discuss.**

## Unbalanced designes

Often your observations will not be clean. This is unfortunate, as it makes the analyses more complicated. But they are still doable.

### Nested ANOVA Example

From Logan Example 11.A - Density-dependent grazing effects of sea urchin on filamentous algea (Andrew and Underwood 1993).

Levels: 
1. Sea urchin density
2. Patches of Reef
3. Quadrats within patches

Get the data and have a look.

```{r}
algae_density <- read.csv(file = "https://mlammens.github.io/Biostats/data/Logan_Examples/Chapter11/Data/andrew.csv")

head(algae_density)
summary(algae_density)
```

Patch and Quad need to be converted to factors!

```{r}
algae_density$PATCH <- factor(algae_density$PATCH)
algae_density$QUAD <- factor(algae_density$QUAD)

summary(algae_density)
```

Much better.

Let's make some visuals.

```{r}
library(ggplot2)

ggplot(data = algae_density) +
  geom_boxplot(aes(x = TREAT, y = ALGAE)) +
  theme_bw()
```

We should also consider this from the perspective to collapsing the quadrats into their mean values.

```{r}
library(dplyr)

algae_density_agg <-
  algae_density %>%
  group_by(TREAT, PATCH) %>%
  summarize(ALGAE_QUAD_MEAN = mean(ALGAE))

ggplot(data = algae_density_agg) +
  geom_boxplot(aes(x = TREAT, y = ALGAE_QUAD_MEAN)) +
  theme_bw()

```

Construct an ANOVA model and look at the results.

```{r}
algae_density_aov <- aov(data = algae_density, ALGAE ~ TREAT + Error(PATCH))

summary(algae_density_aov)
```

* What can we conclude?
* What was the unit of replication?

## Factorial ANOVA

What if we want to look at two, or more, factors where one is not nested in the other?

**See visuals as per Logan, Fig 12.1 and 12.2.**

For formula, see Logan p. 314 and for calculation of F-ratios see Table 12.1.



### Problems from unbalanced design

See Logan Fig 12.3, p. 325. The variance in a model is "explained" sequentially.

### Example - Two factor fixed effects model (Model I)


```{r}
limpets <- read.csv(file = "https://mlammens.github.io/Biostats/data/Logan_Examples/Chapter12/Data/quinn.csv")

head(limpets)
summary(limpets)

limpets$DENSITY <- factor(limpets$DENSITY)
```

Let's get some visuals.

```{r}
ggplot(data = limpets) +
  geom_boxplot(aes(x = DENSITY, y = EGGS)) +
  theme_bw()

ggplot(data = limpets) +
  geom_boxplot(aes(x = SEASON, y = EGGS)) +
  theme_bw()

ggplot(data = limpets) +
  geom_boxplot(aes(x = DENSITY, y = EGGS, fill = SEASON)) +
  theme_bw()

```


And let's run the model.

```{r}
limpets_lm <- lm(data = limpets, formula = EGGS ~ SEASON * DENSITY)
summary(limpets_lm)
anova(limpets_lm)
```

Provided there's time, we'll play with these models.