---
title: "Meeting 5 - Probability and Data Visualization"
author: "Matthew E. Aiello-Lammens"
output:
  html_document:
    toc: yes
---

# Review of Lecture 4

* Any questions about material covered last week?


# Visualizing probability - Venn Diagrams

### Aside: getting and loading R packages

One of the biggest advantages to R is that it has a rich ecosystem of *add-ons*, also known as packages.
These packages are written my members of the R community (for the most part), and freely available on [CRAN](https://cran.r-project.org/). 
As a class, let's navigate to CRAN and have a look at the packages section of the site.

Now, to install and load a package.

```{r, eval=FALSE}
install.packages("VennDiagram")
```

Then to load it into my workspace, and make the functions available:

```{r}
library(VennDiagram)
```


Let's draw a Venn diagram representing the probability of getting a **blue** M&M.
This is written algebraically as $P(Blue)$:

```{r, results='hide'}
## Load VennDiagram package - NB: you will likely have to install this package first
library(VennDiagram)

## Create a Venn diagram
draw.single.venn(area = 0.24, category = "Blue M&M", fill = "blue", alpha = 0.5)
```

Draw a Venn diagram representing the probability of getting a **blue or red** M&M.
This is written algebraically as $P(Blue \cup Red)$:

```{r, results='hide'}
grid.newpage()
draw.pairwise.venn(area1 = 0.24, area2 = 0.13, cross.area = 0, 
                   category = c("Blue M&M", "Red M&M"),
                   fill = c("blue", "red"), alpha = rep(0.5, 2))
```

**IMPORTANT: Note that there is no overlap between these two events. We say that these events are *mutually exclusive*.**

#### Adding Peanut and Almond M&Ms to the mix

Let's imagine the scenario that we add an equal amount of Peanut and Almond M&Ms to our bags of M&Ms. 
Now we can ask a question like, *what is the probability of getting a **blue AND peanut** M&M?*
We can write this algebraically as $P(Blue \cap Peanut)$.

Recall that $P(Blue) = 0.24$, and because all three types of M&Ms are represented equally, the $P(Peanut) = 0.33$.

```{r, results='hide'}
grid.newpage()
draw.pairwise.venn(area1 = 0.24, area2 = 0.33, cross.area = 0.0792, 
                   category = c("Blue M&M", "Peanut M&M"),
                   fill = c("blue", "yellow"), alpha = rep(0.5, 2))
```

**IMPORTANT: Note that in this example, the color and type of M&M are *indpendent* from each other.**

#### Challange

**What is the probability of getting a *blue OR peanut* M&M?** I.e. $P(Blue \cup Peanut)$.

Hint: Note the difference between adding up the area represented in the Venn diagram versus adding the $P(Blue) + P(Peanut)$.

### Conditional probability

***

#### Challange

What is the probability of drawing a blue M&M, CONDITIONAL on it being a peanut M&M?

$$
P(Blue|Peanut) = \frac{P(Blue \cap Peanut)}{P(Peanut)}
$$

Calculate this probability.[^1] 

[^1]:The value should be 0.24. **Why?**

***

#### More general thinking

We can use a bit of algebra and re-write a general form of the equation above as:

$$
P(A|B) P(B) = P(A \cap B)
$$

Now, because $P(A \cap B)$ is identical to $P(B \cap A)$, we can expand this further too:

$$
P(A|B) P(B) = P(A \cap B) = P(B|A) P(A)
$$

And after a little more algebra, we arrive at **Bayes Theorem**:

$$
P(A|B) = \frac{P(B|A) P(A)}{P(B)} = \frac{P(B|A) P(A)}{P(B|A)P(A) + P(B|~A)P(~A)}
$$

***

# Data Visualization and Exploration

The first step to any data analysis is to interrogate your data by calculating some standard statistics and by visualizing it it various ways. 
In future classes we will look more closely at standard statistics.
Today, we'll focus on data visualization.

### Iris dataset

We're going to work with a dataset that comes built in to R, commonly 
called the
[iris dataset](https://stat.ethz.ch/R-manual/R-patched/library/datasets/html/iris.html).
It is also sometimes called Fisher's Iris dataset (but should more appropriately
be called Anderson's Iris dataset). Because it comes pre-packaged with R, we 
can load it into our environment using the `data` function.

```{r}
data( iris )
```

Let's have a look at this dataset

```{r}
head( iris )
tail( iris )
```

The dataset contains measurements of four characteristics of three
different species of Iris (plants!). 

#### `summary` function

Let's begin by using the `summary` function to examine this dataset.
`summary` returns many of the standard statistics.
When doing data exploration, a few things you want to look at are:

* How to the mean and median values within a variable compare?
* Do the min and max values suggest there are outliers?
* Which variables (i.e., columns) are quantitative (numeric) versus categorical (factors or characters)

```{r}
summary(iris)
```

#### Aside: Differences between characters and factors

Factors 'appear' to be similar to characters, but are in fact coded numerically in R. 
Think of **factors** like **categories**. 
Here's a quick example that demonstrates the difference in these two variable types that shows up when using `summary`.

```{r}
## Make a new iris dataset
iris_new <- iris

## Create a new column that treats species as a character, rather than a factor
iris_new$species_char <- as.character(iris_new$Species)

## Run summary command
summary(iris_new)
```


#### Aside: A (very) brief introduction to navigating a `data.frame`

We will be very brief here. 
I recommend checking out 
[this Data Carpentry lesson](http://www.datacarpentry.org/R-ecology/03-data-frames.html) for more information.

* Looking at specific `data.frame` elements. Use the *row* and *column* notation.

Here is the 5th row, 3rd column (Petal.Length).
**Note: We are using square brackets to index the `data.frame` and we *always* use row, column notation.**

```{r}
iris[5, 3]
```

* Looking at an entire column.

Here are two ways to get the `Petal.Length` column.

First, **note: we leave the row part blank, but still add the comma.**

```{r}
iris[ ,3]
```

Second, **use only the variable (column) name. Note the use of the `$` operator**

```{r}
iris$Petal.Length
```

* Looking at specific column entry

This is another way to look at the 5th entry in the `Petal.Length` column.

```{r}
iris$Petal.Length[5]
```

* Looking at all entries for a given row.

Here's all the entries for the 5th row.
**Note: here we leave the column part blank, but still add the comma.**

```{r}
iris[5, ]
```

* Looking at a set of rows and/or columns.

Here's all the entries in the 5th through 10th rows, 1st through 3rd columns.
**Note: we use the `:` operator to look at a range of value.**

```{r}
iris[5:10, 1:3]
```

* For `data.frame`s, if you do not use row, column notation, you will get only the columns back.

```{r}
head(iris[2:3])
```

***

#### Challange

What am I going to get if I execute the command below?

```{r, results='hide'}
head(iris[c("Sepal.Width","Petal.Length")])
```

***

### Visualization using `base` R functions

Let's add a habitat type variable to the iris dataset. We'll use this later. 
*Caveat - I made up these habitat type preferences.*

```{r}
iris_habitat <- data.frame( Species = c( "setosa", "versicolor", "virginica" ),
                            Habitat = c( "forest", "wetland", "meadow" ) )

iris_full <- merge( x = iris, y = iris_habitat, by = "Species" )

head( iris_full )
tail( iris_full )
```

#### Visualizing the measurements of a single variable

Perhaps the most common way to look at data for a single variable is a histogram.
This essentially is a bar plot, where each bar represents the number of times a value falls within a particular *bin*.

Example: let's look at the distribution of `Petal.Length` values

```{r}
hist(iris_full$Petal.Length)
```

***

#### Challange - use `?hist` to determine how to change the number of bins used

Hint: bins are sometimes called breaks

```{r}
?hist
hist(iris_full$Petal.Length, breaks = 30)
```

***

#### Visualizing relationships between two variables

We would like to plot Sepal.Length versus Petal.Length. We'll first do this using
**base R** functions.

```{r plot1}
plot( x = iris_full$Sepal.Length, y = iris_full$Petal.Length )
```

Note, we could save a little typing by using the `with` function. This 
sets up a temporary environment where all of the variables (columns) 
in my dataset are defined individually.

```{r plot1-with}
with( data = iris_full, plot( x = Sepal.Length, y = Petal.Length ) )
```


### Visualization using the `ggplot2` package

Now we're going to introduce a data visualization package called **ggplot2**.
This package is great for producing publication quality graphics, but the syntax
used to create a plot is a little more involved.

#### Aside: Installing and loading packages

First, let's install the `ggplot2` package:

```{r installggplot2, eval=FALSE}
# Only need to do this once
install.packages("ggplot2")
```

Then load it:

```{r loadggplot2, message=FALSE}
library(ggplot2)
```


OK, let's fire up ggplot.

```{r plot1-ggplot}
ggplot() + # First make an empty ggplot canvas. Notice the trailing plus sign
  geom_point( data = iris_full, aes( x = Sepal.Length, y = Petal.Length ) )
```

Let's break down that call to introduce a few key things about ggplot

* ggplot: the initial canvas we're working on
* geom: geometric objects (i.e. the type of plot - histogram, points, line, etc) 
* aes: aesthetic mapping 

**THAT SEEMS SO COMPLICATED!**

It's true, for simple plots, ggplot can be much more complicated
than simply using the base functions. But the power of ggplot lies
in the ability to lay several geometries (geoms) over each other.
Also, each geometry has a rich set of options. For example,
let's say I want to create the plot we just made, but have 
each species represented by a different color.

```{r plot2}
ggplot() + # First make an empty ggplot canvas. Notice the trailing plus sign
  geom_point( data = iris_full, aes( x = Sepal.Length, y = Petal.Length, colour = Species ) )
```

Let's add more information - how about habitat type as well.

```{r}
ggplot() + # First make an empty ggplot canvas. Notice the trailing plus sign
  geom_point( data = iris_full, aes( x = Sepal.Length, y = Petal.Length, colour = Species, shape = Habitat), size = 2.5 ) +
  theme_bw()
```


**facets** - a way to separate data into different subplots

Let's say we wanted different plots for each species. We can do that in ggplot using
`facets`.

```{r plot2-facets}
ggplot( data = iris_full, aes( x = Sepal.Length, y = Petal.Length ) ) + 
  geom_point() +
  facet_grid( Species ~ . )
```

**NOTE**: I moved the `data ...` stuff into the initial `ggplot` call here.

*** 

#### Challange

1. `ggplot2` has many geometries, allowing us to make lot's of different 
types of plots. Let's make two new plots - one **boxplot** of `Petal.Length`, 
with one boxplot for each species. Use `geom_boxplot` for this.

```{r}
ggplot(data = iris_full, aes(x = Species, y = Petal.Length)) +
  geom_boxplot(aes(fill = Species)) +
  theme_bw()  
```

2. Make a histogram of Petal.width. 

```{r}
ggplot() +
  geom_histogram(data = iris_full, aes(x = Petal.Length)) +
  theme_bw()  
```


Use facets to separate the three species.

```{r}
ggplot() +
  geom_histogram(data = iris_full, aes(x = Petal.Length, fill = Species)) +
  facet_grid( Species ~ . ) +
  theme_bw()  
```

3. Second, let's make density plots of Petal.Width. Use `geom_density` and
`colour` to make different colored density lines for iris in each 
habitat type. 
**Note: the area under the curve is equal to 1.**

```{r}
ggplot() +
  geom_density(data = iris_full, aes(x = Petal.Length, colour = Species)) +
  theme_bw()
```

4. Use histogram to plot density instead of counts.

```{r}
ggplot() +
  geom_histogram(data = iris_full, 
                 aes(x = Petal.Length, y = ..density.., fill = Species)) +
  facet_grid( Species ~ . ) +
  geom_density(data = iris_full, aes(x = Petal.Length, colour = Species)) +
  theme_bw()  
```


***

### More `ggplot2` resources

* <http://docs.ggplot2.org/current/>: The official **ggplot2** documentation.
* <http://www.amazon.com/o/ASIN/0387981403/ref=nosim/gettgenedone-20>: The **ggplot2** book, by the developer, Hadley Wickham.
* <https://groups.google.com/forum/#!forum/ggplot2>: The **ggplot2** Google Group (mailing list, discussion forum).
* <https://github.com/swcarpentry/bc/tree/master/intermediate/r/data-visualization>: Intermediate Software Carpentry lesson on data visualization with **ggplot2**.
* <http://learnr.wordpress.com/>: A blog with a good number of posts describing how to reproduce various kind of plots using **ggplot2**.
* <http://stackoverflow.com/questions/tagged/ggplot2>: Thousands of questions and answers tagged with "ggplot2" on Stack Overflow, a programming Q&A site.

***
