---
title: "Meeting 1 - Introduction to Course and R Programming"
author: "Matthew E. Aiello-Lammens"
date: '2018-01-23'
output: 
  html_document: 
    toc: yes
---

# Class expectations

* Why are you taking this class?
* What are you hoping to learn?

See [Syllabus](http://mlammens.github.io/ENS-623-Research-Stats/syllabus/)

# Class goals

* Learn how to develop a research question
* Practice systematic review of the scientific literature
* Become proficient using the R statistical programming language
* Learn how to explore data collected during a research project
* Learn and master the standard statistical analyses used in scientific studies
* Understand the connections between statistical analyses and study design
* Become familiar with resources to explore advanced analysis methods

# Class format

* Some lecture - mostly about concepts
* Lots of **live coding** examples - you should follow along, but don't have to
* Challenges - some will be short, some long. Some you work on alone, some with a partner or in groups

# Hypothesis testing and the scientific method

### What is a theory?

A proposed explanation for some event or process that is based on the available evidence related to that event.
We also call this a **model**. *Models don't necessarily mean **math**.*

### What is a hypothesis?

A prediction based on our model (or theory). 
For a hypothesis to be considered **scientific** it must be **falsifable**.

## Scientific method

Traditional view:

1. Develop a hypothesis 
2. Make a prediction based on hypothesis 
3. Do an experiment to test the prediction
4. Analyze the data from our experiment
5. Based on results, make an inference about our hypothesis - do the data support or reject our prediction?
6. Repeat as necessary to acquire knowledge

Checkout [Understanding Science](http://undsci.berkeley.edu/article/0_0_0/howscienceworks_02)

### Analysis is iterative

The process of analyzing data is often iterative. 
That means, we may start looking at our data, applying analysis techniques that might *not* be the techniques we use in our final report (i.e., our reports and manuscripts). 
Don't be afraid to start exploring your data, and applying analyses, before you think you're ready.

# Introduction to R and RStudio

## Why do I have to learn to program in R?

In short, it's "industry standard". 
But the longer answer has to do with replicability of data analyses.
Documenting your analysis in R, using R scripts or Rmd files allows you to re-run your analysis whenever the need arises, and to share your analysis workflow with others.

## Getting started

### Overall layout

* Four panels, each has valuable information

### File management

* Setup a `Biostatistics` directory
* Make a `data` directory
* Make a `scripts` directory

### Getting help

* **Help** panel (lower right corner)
* `help.search`

```{r}
help.search("bar plot")
```

**Challenge**

Use the `help.search` function to search for something in statistics that you think should be in R? Did you find anything?

* I know my function - just give me the details - `?barplot`


## R as calculator

We can use R just like any other calculator.

```{r}
3 + 5
```

There's internal control for order of operations (Please Excuse My Dear Aunt Sally)

```{r}
(3 * 5) + 7
```

```{r}
3 * 5 + 7
```

**Challenge**

Write an example where adding parentheses matters.

There are a ton of internal functions, and a lot of add-ons.

```{r}
sqrt(4)

abs(-5)

sqrt(-5)
```

## R script file

Use a script file for your work. It's easier to go back to and easy to document.

**Important:** within an R file, you can use the # sign to add comments. Anything written after the # is *not* interpreted when you run the code.

**Challenge**

Create a new R script file in your `scripts` directory.

### Basic file managment in R

```{r}
# What working directory am I in?
getwd()

# Move to a different director?
setwd(".")
```

#### Things to cover

* Navigating the file path
* Tab completion of file paths
* Tab completion of R commands

**Challenge**

* Try to auto-complete `fil`, what do you find?
* Use the brief help menu that comes up to find a function that starts with `file`, and describe what you think it does.

## Rmd file

Use this to integrate text and R code into the same document.
**I will expect most of your homework assignments as an Rmd file.**

*Practice with Rmd file*

## Variables and objects

There are several basic types of data structures in R.

* **VECTORS**: One-dimensional arrays of numbers, character strings, or logical values (T/F)
* **FACTORS**: One-dimensional arrays of factors (Stop - Let's discuss factors)
* **DATA FRAMES**: Data tables in which the various columns may be of different type
* **MATRICES**: In R, matrices can only be 2-dimensional, anything higher dimension is called an array (see below). Matrix elements are numerical; some functions, like the transpose function t(), only work on matrices
* **ARRAYS**: higher dimensional matrices are called arrays in R
* **LISTS**: lists can contain any type of object as list elements. You can have a list of numbers, a list of matrices, a list of characters, etc., or any combination of the above.

### Functions that are useful for understanding the different types of data structures

```
str()
class()
```

## Practice with variables

Define a variable

```{r}
my_var <- 8
```

And another

```{r}
my_var2 <- 10
```

Work with vars

```{r}
my_var + my_var2
```

Make a new variable

```{r}
my_var_tot <- my_var + my_var2
```

**Challenge**

Change the value of my_var2

```{r}
my_var2 <- 3
```

What is the value of `my_var_tot` now?

### Make a vector

Combining values into a vector

```{r}
# Vector of variables
my_vect <- c(my_var, my_var2)

# Numeric vector
v1 <- c(10, 2, 8, 7, 11, 15)

# Char vector
pets <- c("cat", "dog", "rabbit", "pig")
```

Making a vector of numbers in sequence

```{r}
v2 <- 1:10
v3 <- seq(from = 1, to = 10)
```

**Challenge**

1. Look up the help for the `seq` function, and use this to make a vector from 1 to 100, by steps of 5.
2. Come up with a way that you would use the `length.out` argument.


