---
title: "Meeting 2 - R Programming Cont'd"
author: "Matthew E. Aiello-Lammens"
output:
  html_document:
    toc: yes
---

# Asking research questions

* Brief background of Prof. Aiello-Lammens' graduate research

> *... it is not realistic to expect yourself to sit at your desk and conjure up the perfect study that will revolutionize the field.* -Karban et al. 2014, p. 5

* Coming up with your question - how did Karban et al. 2014, Chapter 1 make you think about this?

> *Be willing to hang with the dumb ideas that you will inevitably come up with, because the really great ideas stand on the shoulders of the dumb ones.* -Karban et al. 2014, p. 14

* How to approach "the literature"
    - Peer-review is king, but not all journals are equal
    - Follow the paper trail

* Grants and external funding - concentrating your thoughts

# Introduction to R - continued

## Exploring variable elements

You can get specific elements from vectors and other data structures

* Introduction to the square brackets `[]`

```{r}
pets <- c("cat", "dog", "rabbit", "pig", "snake")
pets[1]
```

* Getting a number of elements, in sequence

```{r}
pets[3:4]
```

* Getting a number of elements, not in sequence

```{r}
pets[c(1,4)]
```


## Working with matrices

**Review** - Why might we want 2D data?

Let's make a matrix

**Challenge**

With the people next to you, break down this function, and describe each argument.
What is the final product?

```{r}
my_mat <- matrix(data = runif(50), nrow = 10, byrow = TRUE)
```

What does it mean to fill `byrow`?

```{r}
matrix(data = 1:9, nrow = 3, byrow = TRUE)
```

Versus

```{r}
matrix(data = 1:9, nrow = 3, byrow = FALSE)
```

**Challenge**

What is the default value for `byrow`?

### Indexing matrices

Indexing happens by **row**, **column** notation.

```{r}
my_mat <- matrix(data = 1:50, nrow = 10, byrow = TRUE)

my_mat[1,1]
my_mat[1,2]
my_mat[2,1]

my_mat[1:4, 1:3]

my_mat[c(1,3,5), ]
my_mat[ ,c(1,3,5)]
```

## Combinining internal functions with matrices

Make a "random" matrix (that isn't random in this case because of the `set.seed` function)

```{r}
set.seed(1)
mat1 <- matrix(data = runif(50), nrow = 10, byrow = TRUE)
```

Calculate the mean of all of the data

```{r}
mean(mat1)
```

Calculate the standard deviation of all of the data

```{r}
sd(mat1)
```

Calculate row means and column means

```{r}
rowMeans(mat1)
colMeans(mat1)
```

Introduce the `apply` function

```{r}
apply(mat1, MARGIN = 1, mean)
```

## User Defined Functions

What is a function?

Why write a function?

**Example**

*Note:* the example below is borrowed from the [Software Carpentry](https://software-carpentry.org/) Introduction to Programming materials.

Convert from Fahrenheit to Kelvin

```{r}
fahr_to_kelvin <- function(temp) {
  kelvin <- ((temp - 32) * (5 / 9)) + 273.15
  return(kelvin)
}
```

Convert from Kelvin to Celsius

```{r}
kelvin_to_celsius <- function(temp) {
  celsius <- temp - 273.15
  return(celsius)
}
```

**Challenge**

Write a function to convert from Fahrenheit to Celsius.


## `for` loops

How do I do the same thing many times?

### Generic for loop

```
for (variable in collection) {
  do things with variable
}
```

Let's get more specific. Say we took a bunch of measurements of temperature in Fahrenheit, but want to convert them. How might we do it?

#### Make our temperature data set

```{r}
set.seed(8)
temp_data <- runif(n = 20, min = -5, max = 5) + 45
```

**Challenge**

What did we just do?

#### Use our `fahr_to_kelvin` function on each element

Iteration 1.

```{r}
for( x in temp_data){
  fahr_to_kelvin(x)
}
```

Iteration 2.

```{r}
for( x in temp_data){
  print(fahr_to_kelvin(x))
}
```

Iteration 3.

```{r}
for( x in 1:length(temp_data)){
  print(fahr_to_kelvin(temp_data[x]))
}
```

Iteration 4.

```{r}
temp_data_kelvin <- vector()
for( x in 1:length(temp_data)){
  temp_data_kelvin[x] <- fahr_to_kelvin(temp_data[x])
}

print(temp_data_kelvin)
```



## Conditionals

We can use conditional statements to control the flow of our code, and to "make choices" as it progresses.

### `if` and `else`

The `if` and `else` statements are key to making choices in your code.
Before understanding `if`/`else` statements, we need to review **booleans** - i.e., `TRUE` and `FALSE`

#### Aside: `TRUE` and `FALSE` values

```{r}
TRUE
```

```{r}
20 == 20
```

```{r}
20 > 40
```

A `!` sign can be used as a logical negation

```{r}
!(20 > 40)
```

There are many logical operators to consider.

***

**Challenge** Describe what each fo the following operators are doing.

```{r}
x <- TRUE
y <- FALSE
```

```{r}
x & y
```

```{r}
x | y
```

```{r}
xy <- c(x,y)
```

```{r}
any(xy)
```

```{r}
all(xy)
```

***

Now that we know a bit about booleans, let's get into `if`/`else` statements.

Essential, and `if` is a conditional that says, "do this thing after the `if` statement, ***if*** the conditional was TRUE."

Here's a simple example (taken from Software Carpentry's lessons)

```{r}
num <- 37
if (num > 100) {
  print("greater")
} else {
  print("not greater")
}
```

**Challenge** 

Re-define `num` so you get the other option.

We don't need an `else` statement for this to work -

```{r}
num <- 37
if (num > 100) {
  print("The number was greater than 100")
}

```

We can also write a "cascade" of `if`/`else` statements

```{r}
if (num > 0) {
  return(1)
} else if (num == 0) {
  return(0)
} else {
  return(-1)
}
```

**Challenge**

Make the above into a function, that takes in any value, and returns whether it is positive, negative, or equal to 0.



