---
title: "Meeting 4 - Introduction to Probability"
author: "Matthew E. Aiello-Lammens"
output:
  html_document:
    toc: yes
---

# Probability

> *We usually talk about probabilities in terms of events; the probability of event A occurring is written P(A). Probabilities can be between zero and one; if P(A) equals zero, then the event is impossible; if P(A) equals one, then the event is certain.* - Quinn and Keough 2002

# Frequentist statistics 

The relative frequency of an event over the long term, after many trials.

E.g., throwing a coin. What's the probability of a 'heads'? After one trial, two trials, ..., 1000 trials.

```{r}
(head <- runif(1) > 0.5)
```


## Bayesian statistics

For now, all I will say is that there are other perspectives in statistics. We'll cover a bit of Bayes at another time.

***

# Probability through M&Ms

Open your bag(s) and count how many M&Ms you got. Also count how many of each color.

Make a new variable called `tot_mm` and assign it the number of MMs in your bag.

```{r}
tot_mm <- 60
```

## `mean` number of MMs in a bag

Calculate the mean number of MMs in a bag.

```{r}
tot_mm_pop_sample <- c(60, 58, 57, 63, 61, 60, 59)
```

Mean:

$$
\bar{x} = \frac{\sum_{i=1}^n{x_i}}{n}
$$

Using the `sum` function.

```{r}
(mean_mm_1 <- sum(tot_mm_pop_sample) / length(tot_mm_pop_sample))
```

Using the `mean` function.

```{r}
(mean_mm_2 <- mean(tot_mm_pop_sample))
```

## Quantify the variation in the number of MMs in a bag

* `var` function
* `sd` function

**What do we mean by variation?**

In general, we mean something along the lines of 'the amount of variation around some mean value'.

**How might we quantify this?**

* Add up the distances from the mean. 

First calculate distances from the mean for each bag.

```{r}
(dist_from_mean <- tot_mm_pop_sample - mean_mm_1)
```

Next calculate the total distances. We could also call this the **sums of distances**

```{r}
(sum_dist_from_mean <- sum(dist_from_mean))
```

What information does this give us, and why might it not be useful?

* Add up the *absolute* distances form the mean. 

First calculate the *absolute* distances from the mean.

```{r}
(absdist_from_mean <- abs(tot_mm_pop_sample - mean_mm_1))
```

Next add up the total *absolute* distances from the mean.

```{r}
(sum_absdist_from_mean <- sum(absdist_from_mean))
```

* Add up the *squared* distances from the mean.

While the absolute distance from the mean does give us a reasonable measure of variability, because of specific mathematical properties, it's more convenient to work with *squared* distances from the mean, leading to the measures of **variance** and **standard deviation**.

**Challenge**

Calculate *squared* distances from the mean, and sum them to determine the total squared deviation.

**Standard deviation** can be thought of as the average deviation from the mean.

**Challenge**

1. Calcualte the **variance** of the number of M&Ms in a bag, considering all of the bags in the class. Do this without using the `var` function.

2. Calculate the **standard deviation** of the number of M&Ms in a bag. Do this without using the `sd` function.

## Probability and M&Ms

Without looking, you chose one M&M. What colors could you have chosen?

The **set** of brown, yellow, green, red, orange, blue is the **sample space**.

Your selection of a single M&M is called an **event**.


**Challenge**

* What's the probability of getting a "color" M&M in **your** bag?

P(brown), P(yellow), etc.

### Putting our data together into a `data.frame`

Here we will collate the classes data into a single data set.

**Challenge**

What is the mean probability of getting each color across each bag?

Tools to use:

* `data.frames`
* `apply`


## Probability of events

* What's the P(green OR blue OR red) in **your** bag? 

P(green) + P(blue) + P(red)

* What is the P(NOT green) in **your** bag?

$$
P(\sim Green) = 1 - P(Green) = P(G)^c
$$


### Sampling with replacement

I draw one M&M from my bag, put it back, then draw another.

* What's the P(green and then blue)?

<!---
P(green) * P(blue) 
--->

* What's the P(green or blue)?

<!---
P(green) + P(blue)
--->

### Sampling without replacement

* What is the probability of getting green, then blue, *without replacing your first draw*?

**Challenge**

What is the sample space when drawing two M&Ms?


## Draw a random bag of M&Ms using the company stated frequencies/probabilities for each color

**Challenge**

How would you create a random bag of M&Ms, assuming that each M&M has an equal probability of being in any given bag?



Mars claims that percentages of each color M&Ms are slightly different.

```{r}
## Colors as a vector
mm_colors <- c("blue", "brown", "green", "orange", "red", "yellow")
mm_probs <- c(.24, .14, .16, .20, .13, .14)

## I want to "sample" a bag of MMs
new_bag <- sample(x = mm_colors, size = 55, replace = TRUE, prob = mm_probs)
table(new_bag)

```


# Visualizing probability - Venn Diagrams

### Aside: getting and loading R packages

One of the biggest advantages to R is that it has a rich ecosystem of *add-ons*, also known as packages.
These packages are written my members of the R community (for the most part), and freely available on [CRAN](https://cran.r-project.org/). 
As a class, let's navigate to CRAN and have a look at the packages section of the site.

Now, to install and load a package.

```{r, eval=FALSE}
install.packages("VennDiagram")
```

Then to load it into my workspace, and make the functions available:

```{r}
library(VennDiagram)
```



Let's draw a Venn diagram representing the probability of getting a **blue** M&M.
This is written algebraically as $P(Blue)$:

```{r, results='hide'}
## Load VennDiagram package - NB: you will likely have to install this package first
library(VennDiagram)

## Create a Venn diagram
draw.single.venn(area = 0.24, category = "Blue M&M", fill = "blue", alpha = 0.5)
```

Draw a Venn diagram representing the probability of getting a **blue or red** M&M.
This is written algebraically as $P(Blue \cup Red)$:

```{r, results='hide'}
grid.newpage()
draw.pairwise.venn(area1 = 0.24, area2 = 0.13, cross.area = 0, 
                   category = c("Blue M&M", "Red M&M"),
                   fill = c("blue", "red"), alpha = rep(0.5, 2))
```

**IMPORTANT: Note that there is no overlap between these two events. We say that these events are *mutually exclusive*.**

#### Adding Peanut and Almond M&Ms to the mix

Let's imagine the scenario that we add an equal amount of Peanut and Almond M&Ms to our bags of M&Ms. 
Now we can ask a question like, *what is the probability of getting a **blue AND peanut** M&M?*
We can write this algebraically as $P(Blue \cap Peanut)$.

Recall that $P(Blue) = 0.24$, and because all three types of M&Ms are represented equally, the $P(Peanut) = 0.33$.

```{r, results='hide'}
grid.newpage()
draw.pairwise.venn(area1 = 0.24, area2 = 0.33, cross.area = 0.0792, 
                   category = c("Blue M&M", "Peanut M&M"),
                   fill = c("blue", "yellow"), alpha = rep(0.5, 2))
```

**IMPORTANT: Note that in this example, the color and type of M&M are *indpendent* from each other.**

#### Challange

**What is the probability of getting a *blue OR peanut* M&M?** I.e. $P(Blue \cup Peanut)$.

Hint: Note the difference between adding up the area represented in the Venn diagram versus adding the $P(Blue) + P(Peanut)$.

### Conditional probability

***

#### Challange

What is the probability of drawing a blue M&M, CONDITIONAL on it being a peanut M&M?

$$
P(Blue|Peanut) = \frac{P(Blue \cap Peanut)}{P(Peanut)}
$$

Calculate this probability.[^1] 

[^1]:The value should be 0.24. **Why?**

***

#### More general thinking

We can use a bit of algebra and re-write a general form of the equation above as:

$$
P(A|B) P(B) = P(A \cap B)
$$

Now, because $P(A \cap B)$ is identical to $P(B \cap A)$, we can expand this further too:

$$
P(A|B) P(B) = P(A \cap B) = P(B|A) P(A)
$$

And after a little more algebra, we arrive at **Bayes Theorem**:

$$
P(A|B) = \frac{P(B|A) P(A)}{P(B)} = \frac{P(B|A) P(A)}{P(B|A)P(A) + P(B|~A)P(~A)}
$$

***

