---
title: "Lecture 1"
author: "Matthew E. Aiello-Lammens"
date: '2016-01-25'
output: html_document
---

# Class expectations

* Why are you taking this class?
* What are you hoping to learn by taking this class?

See [Syllabus](http://mlammens.github.io/Biostats/syllabus/)


# Hypothesis testing and the scientific method

### What is a theory?

A proposed explanation for some event or process that is based the available evidence related to that event.
We also call this a **model**. *Models are not necessarily equivelant to math.*

### What is a hypothesis?

A prediction based on our model or theory. 
For a hypothesis to be considered **scientific** it much be **falsifable**.

## Scientific method

Traditional view:

1. Develop a hypothesis 
2. Make a prediction based on hypothesis 
3. Do an experiment to test the prediction
4. Analyze the data from our experiment
5. Based on results, make an inference about our hypothesis - do the data support or reject our prediction?
6. Repeat as necessary to acquire knowledge

Checkout [Understanding Science](http://undsci.berkeley.edu/article/0_0_0/howscienceworks_02)

# Introduction to R and RStudio

## Getting started

### Overall layout

* Four panels, each has valuable information

### File management

* Setup a `Biostatistics` directory
* Make a `data` directory
* Make a `scripts` directory

### Getting help

* **Help** panel (lower right corner)
* `help.search`

```{r}
help.search("bar plot")
```

**Challange**

Use the `help.search` function to search for something in statistics that you think should be in R? Did you find anything?

* I know my function - just give me the details - `?barplot`


## R as calculator

We can use R just like any other calculator.

```{r}
3 + 5
```

There's internal control for order of operations (Please Excuse My Dear Aunt Sally)

```{r}
(3 * 5) + 7
```

```{r}
3 * 5 + 7
```

**Challange**

Writen an example where adding parantheses matters.

There are a ton of internal functions, and a lot of add-ons.

```{r}
sqrt(4)

abs(-5)

sqrt(-5)
```

## R script file

Use a script file for your work. It's easier to go back to and easy to document.

## Variables and objects

There are several basic types of data structures in R.

* **VECTORS**: One-dimensional arrays of numbers, character strings, or logical values (T/F)
* **FACTORS**: One-dimensional arrays of factors (Stop - Let's discuss factors)
* **DATA FRAMES**: Data tables in which the various columns may be of different type
* **MATRICES**: In R, matrices can only be 2-dimensional, anything higher dimension is called an array (see below). Matrix elements are numerical; some functions, like the transpose function t(), only work on matrices
* **ARRAYS**: higher dimensional matrices are called arrays in R
* **LISTS**: lists can contain any type of object as list elements. You can have a list of numbers, a list of matrices, a list of characters, etc., or any combination of the above.

Functions that are useful for understanding the different types of data structures

```
str()
class()
```

# Probability

### Frequentist perspective 

The relative frequency of an event over the long term, after many trials.

E.g., throwing a coin. What's the probablity of a 'heads'? After one trial, two trials, ..., 1000 trials.

```{r}
(head <- runif(1) > 0.5)
```


### Bayes perspective

Stay tuned.

***


## M&Ms

Open your bag and count how many M&Ms you got. Also count how many of each color.

### Assignment operator

Make a new variable (object) called `tot_mm` and assign it the number of MMs in your bag.

```{r}
tot_mm <- 60
```

### `mean` number of MMs in a bag

* Vector
* `sum` function
* `mean` function

Calculate the mean number of MMs in a bag.

```{r}
tot_mm_pop_sample <- c(60, 58, 57, 63, 61, 60, 59)
```

Mean:

$$
\bar{x} = \frac{\sum_{i=1}^n{x_i}}{n}
$$

Using the `sum` function.

```{r}
(mean_mm_1 <- sum(tot_mm_pop_sample) / length(tot_mm_pop_sample))
```

Using the `mean` function.

```{r}
mean_mm_2 <- mean(tot_mm_pop_sample)
```

### Quantify the variation in the number of MMs in a bag

* `var` function
* `sd` function

What do we mean by variation? 
In general, we mean something along the lines of 'the amount of variation around some mean value'.

**How might we quantify this?**

* Add up the distances from the mean. 

First calculate distances from the mean for each bag.

```{r}
(dist_from_mean <- tot_mm_pop_sample - mean_mm_1)
```

Next calculate the total distances.

```{r}
(sum_dist_from_mean <- sum(dist_from_mean))
```

What information does this give us, and why might it not be useful?

* Add up the *absolute* distances form the mean. 

First calculate the *absolute* distances from the mean.

```{r}
(absdist_from_mean <- abs(tot_mm_pop_sample - mean_mm_1))
```

Next add up the total *absolute* distances from the mean.

```{r}
(sum_absdist_from_mean <- sum(absdist_from_mean))
```

* Add up the *squared* distances from the mean.

While the absolute distance from the mean does give us a reasonable measure of variability, because of specific mathematical properties, it's more convenient to work with *squared* distances from the mean, leading to the measures of **variance** and **standard deviation**.

**Challange**

Calculate *squared* distances from the mean, and sum them to determine the total squared deviation.

**Challange**

Can you find an R function to do this in one step?


## Probability and M&Ms

Without looking, you chose one M&M. What colors could you have chosen?

The **set** of brown, yellow, green, red, orange, blue is the **sample space**.

Your selection of a single M&M is called an **event**.


**Challange**

* What's the probability of getting a "color" M&M in **your** bag?

P(brown), P(yellow), etc.

**Challange**

What is the mean probablity of getting each color across each bag?

Tools to use:

* `data.frames`
* `apply`


## Venn diagrams

* What's the P(green OR blue OR red) in **your** bag? 

P(green) + P(blue) + P(red)

* What is the P(NOT green) in **your** bag?

$$
P(NOT Green) = 1 - P(Green) = P(G)^c
$$


### Sampling with replacement

* What's the P(green and then blue), first putting back your first draw?

Intersection: P(green intersection blue) = P(green) * P(blue) 

* What's the P(green or blue)?

Union: P(green) + P(blue) - P(green and blue)


### Sampling without replacement

* What is the probability of getting green, then blue, *without replacing your first draw*?

**Challange**

What is the sample space for drawing two M&Ms?

Well work further on similar problems in future classes.

### Conditional probability

* What's the probability of drawing a red M&M, CONDITIONAL on having a bag with 60 M&Ms?

### Draw a random bag of M&Ms using the company stated frequencies/probabilities for each color

```{r}
## Colors as a vector
mm_colors <- c("blue", "brown", "green", "orange", "red", "yellow")
mm_probs <- c(.24, .14, .16, .20, .13, .14)

## I want to "sample" a bag of MMs
new_bag <- sample(x = mm_colors, size = 55, replace = TRUE, prob = mm_probs)
table(new_bag)

```


