---
title: "Lecture 7 - Linear Models, Correlation, and Regression"
author: "Matthew E. Aiello-Lammens"
output: 
  html_document: 
    toc: yes
---

# Logistics

* Solution sets for problem sets 1-3 will be posted on the Open Science Framework site on 3/21. This will also be the last day to submit any past assignments.
* The take-home midterm will be distributed tomorrow afternoon.
* The midterm is due Monday, 3/28.

# Relationships among the distributions

![distribution relationships](/Volumes/Garage/Professional-Annex/SBU-Grad-School/SBU-Teaching/Biometry-BEE-552-SP13/Figure adapted from Leemis (1986).pdf) 


### Simple example

Effects of starvation and humidity on water loss in the confused flour beetle.
Here, looking at the relationship between humidity and weight loss

```{r}
flr_beetle <- read.csv(file = "https://mlammens.github.io/Biostats/data/Logan_Examples/Chapter8/Data/nelson.csv")
flr_beetle
```

Plot these data

```{r}
ggplot() +
  geom_point(data = flr_beetle, aes( x = HUMIDITY, y = WEIGHTLOSS)) +
  theme_bw()
```

Run a linear regression

```{r}
flr_beetle_lm <- lm(data = flr_beetle, WEIGHTLOSS ~ HUMIDITY)

flr_beetle_lm

summary(flr_beetle_lm)
```

Plot these data, with `lm` fit

```{r}
ggplot(data = flr_beetle, aes( x = HUMIDITY, y = WEIGHTLOSS)) +
  geom_point() +
  stat_smooth(method = "lm") +
  theme_bw()
```

