---
title: "Lecture 2"
author: "Matthew E. Aiello-Lammens"
output:
  html_document:
    toc: yes
---


## Exploring variable elements

You can get specific elements from vectors and other data structures

* Introduction to the square brackets `[]`

```{r}
pets[1]
```

* Getting a number of elements, in sequence

```{r}
pets[3:4]
```

* Getting a number of elements, not in sequence

```{r}
pets[c(1,4)]
```

**Challenge**

Use what we just learned to reverse the order of the `pets` vector.

### Working with matrices

Why might we want 2D data?

Let's make a matrix

**Challenge**

With the people next to you, break down this function, and describe each argument.
What is the final product?

```{r}
my_mat <- matrix(data = runif(50), nrow = 10, byrow = TRUE)
```

What does it mean to fill `byrow`?

```{r}
matrix(data = 1:9, nrow = 3, byrow = TRUE)
```

Versus

```{r}
matrix(data = 1:9, nrow = 3, byrow = FALSE)
```

**Challenge**

What is the default value for `byrow`?

### Indexing matrices

Indexing happens by **row**, **column** notation.

```{r}
my_mat <- matrix(data = 1:50, nrow = 10, byrow = TRUE)

my_mat[1,1]
my_mat[1,2]
my_mat[2,1]

my_mat[1:4, 1:3]

my_mat[c(1,3,5), ]
my_mat[ ,c(1,3,5)]
```

## Combinining internal functions with matrices

Make a "random" matrix (that isn't random in this case)

```{r}
set.seed(1)
mat1 <- matrix(data = runif(50), nrow = 10, byrow = TRUE)
```

Calculate the mean of all of the data

```{r}
mean(mat1)
```

Calculate the standard deviation of all of the data

```{r}
sd(mat1)
```

Calculate row means and column means

```{r}
rowMeans(mat1)
colMeans(mat1)
```

Introduce the `apply` function

```{r}
apply(mat1, MARGIN = 1, mean)
```

## Functions

What is a function?

Why write a function?

**Example**

*Note:* the example below is borrowed from the [Software Carpentry](https://software-carpentry.org/) Introduction to Programming materials.

Convert from Fahrenheit to Kelvin

```{r}
fahr_to_kelvin <- function(temp) {
  kelvin <- ((temp - 32) * (5 / 9)) + 273.15
  return(kelvin)
}
```

Convert from Kelvin to Celsius

```{r}
kelvin_to_celsius <- function(temp) {
  celsius <- temp - 273.15
  return(celsius)
}
```

**Challenge**

Write a function to convert from Fahrenheit to Celsius.


## `for` loops

How do I do the same thing many times?

### Generic for loop

```
for (variable in collection) {
  do things with variable
}
```

Let's get more specific. Say we took a bunch of measurements of temperature in Fahrenheit, but want to convert them. How might we do it?

#### Make our temperature data set

```{r}
set.seed(8)
temp_data <- runif(n = 20, min = -5, max = 5) + 45
```

**Challenge**

What did we just do?

#### Use our `fahr_to_kelvin` function on each element

Iteration 1.

```{r}
for( x in temp_data){
  fahr_to_kelvin(x)
}
```

Iteration 2.

```{r}
for( x in temp_data){
  print(fahr_to_kelvin(x))
}
```

Iteration 3.

```{r}
for( x in 1:length(temp_data)){
  print(fahr_to_kelvin(temp_data[x]))
}
```

Iteration 4.

```{r}
temp_data_kelvin <- vector()
for( x in 1:length(temp_data)){
  temp_data_kelvin[x] <- fahr_to_kelvin(temp_data[x])
}

print(temp_data_kelvin)
```

**Final challenge**

Use a `for` loop to write a script to calculate the mean of a set of numbers.

Recall that the mean is:

$$
\bar{x} = \frac{\sum_{i=1}^n{x_i}}{n}
$$








# Start 2016 material





# Week 2 - Probability and Data Exploration

## Review from Week 1

### Probability and Venn diagrams

Recall our bag's of M&Ms. 
I told you that Mars claims that percentages of each color M&M should be:

```{r}
## Colors as a vector
mm_colors <- c("blue", "brown", "green", "orange", "red", "yellow")
## Proportion/probability of each color
mm_probs <- c(.24, .14, .16, .20, .13, .14)
```

Let's draw a Venn diagram representing the probability of getting a **blue** M&M.
This is written algebraically as $P(Blue)$:

```{r, results='hide'}
## Load VennDiagram package - NB: you will likely have to install this package first
library(VennDiagram)

## Create a Venn diagram
draw.single.venn(area = 0.24, category = "Blue M&M", fill = "blue", alpha = 0.5)
```

Draw a Venn diagram representing the probability of getting a **blue or red** M&M.
This is written algebraically as $P(Blue \cup Red)$:

```{r, results='hide'}
grid.newpage()
draw.pairwise.venn(area1 = 0.24, area2 = 0.13, cross.area = 0, 
                   category = c("Blue M&M", "Red M&M"),
                   fill = c("blue", "red"), alpha = rep(0.5, 2))
```

**IMPORTANT: Note that there is no overlap between these two events. We say that these events are *mutually exclusive*.**

#### Adding Peanut and Almond M&Ms to the mix

Let's imagine the scenario that we add an equal amount of Peanut and Almond M&Ms to our bags of M&Ms. 
Now we can ask a question like, *what is the probability of getting a **blue AND peanut** M&M?*
We can write this algebraically as $P(Blue \cap Peanut)$.

Recall that $P(Blue) = 0.24$, and because all three types of M&Ms are represented equally, the $P(Peanut) = 0.33$.

```{r, results='hide'}
grid.newpage()
draw.pairwise.venn(area1 = 0.24, area2 = 0.33, cross.area = 0.0792, 
                   category = c("Blue M&M", "Peanut M&M"),
                   fill = c("blue", "yellow"), alpha = rep(0.5, 2))
```

**IMPORTANT: Note that in this example, the color and type of M&M are *indpendent* from each other.**

#### Challange

**What is the probability of getting a *blue OR peanut* M&M?** I.e. $P(Blue \cup Peanut)$.

Hint: Note the different between adding up the area represented in the Venn diagram versus adding the $P(Blue) + P(Peanut)$.

### Conditional probability

***

#### Challange

What is the probability of drawing a blue M&M, CONDITIONAL on it being a peanut M&M?

$$
P(Blue|Peanut) = \frac{P(Blue \cap Peanut)}{P(Peanut)}
$$

Calculate this probability. The value should be 0.24. **Why?**

***

#### More general thinking

We can use a bit of algebra and re-write a general form of the equation above as:

$$
P(A|B) P(B) = P(A \cap B)
$$

Now, because $P(A \cap B)$ is identical to $P(B \cap A)$, we can expand this further too:

$$
P(A|B) P(B) = P(A \cap B) = P(B|A) P(A)
$$

And after a little more algebra, we arrive at **Bayes Theorem**:

$$
P(A|B) = \frac{P(B|A) P(A)}{P(B)} = \frac{P(B|A) P(A)}{P(B|A)P(A) + P(B|~A)P(~A)}
$$

***

## `if`/`else` statements and `for` loops

In this class, and in ecological analysis more generally, you are likely to encounter situations where you want to perform some calculation multiple times. 
A `for` loop is standard programming method used to do this.

### Example - 100 coin flips

Last week we simulated coin flips using both a fair and unfair coin. 
We used the `sample` function to do this, which is a pretty efficient method.
What we're going to use here is much less efficient, but instructive.

First, let's deal with a single coin flip of a fair coin.
I'll do this by drawing a single number from a
[standard uniform](https://en.wikipedia.org/wiki/Uniform_distribution_(continuous)) 
distribution. 
If the number is greater than 0.5, I'll call the "flip" a heads. 
Otherwise I'll call it a "tails".

#### Single coin flip using `if`/`else`

```{r}
coin_flip_number <- runif(n = 1)

if(coin_flip_number > 0.5){
  coin_flip <- "heads"
  } else {
    coin_flip <- "tails"
  }

coin_flip
```

***

#### Challange - Single coin flip using `ifelse`

Use the `ifelse` function to make this a bit cleaner. 
**Hint: start by looking at the help for the `ifelse` function**

```{r}
coin_flip <- ifelse(runif(1) > 0.5, yes = "heads", no = "tails")
```

***

#### Use a `for` loop to simulate 100 coin flips

Let's dive in to a `for` loop. For more information, check out this lesson on loops from
[Software Carpentry](http://swcarpentry.github.io/r-novice-inflammation/03-loops-R.html).

**NOTE: Results are hidden for presentation here.**

```{r, results='hide'}
## Set a variable for the number of "repetitions" you want to do - here 100
rep <- 100

## Initiate for loop
for( ind in 1:rep ){
  
  ## Print out a statement telling you what repetition the loop is on
  print(paste0("Doing repetition: ", ind))
  
  ## Simulate the coin flip
  coin_flip <- ifelse(runif(1) > 0.5, yes = "heads", no = "tails")
  
  ## Print the result of the flip
  print(coin_flip)

}
```

This is good, but if we look at `coin_flip` we see that it only saves the last simulated value, not all 100. 
We need to add a few more things to our loop.

**NOTE: Results are hidden for presentation here.**

```{r, results='hide'}
## Set a variable for the number of "repetitions" you want to do - here 100
rep <- 100

## NEW!!! Make an empty vector to save the results of our coin flips
all_flips <- c()

## Initiate for loop
for( ind in 1:rep ){
  
  ## Print out a statement telling you what repetition the loop is on
  print(paste0("Doing repetition: ", ind))
  
  ## Simulate the coin flip
  coin_flip <- ifelse(runif(1) > 0.5, yes = "heads", no = "tails")
  
  ## Print the result of the flip
  print(coin_flip)
  
  ## NEW!!! Add the coin flip to the `all_flips` vector
  all_flips[ind] <- coin_flip

}
```

Okay, now we can look at the results of our simulated coin flips.

```{r}
table(all_flips)

## or heads = 
sum(all_flips=="heads")

## and tails = 
sum(all_flips=="tails")

```


***

### Major Challange (15 minutes): How likely was my bag of M&Ms?

Write a short R script to simulate the combinations of colors that would have been possible from your bag of M&Ms assuming the company-stated color distribution (24% blue, 14% brown, 16% green, 20% orange, 13% red, 14% yellow) and use this script to calculate the probability of obtaining the combination in your bag. 

Needed:
* `for` loop
* `if` statement
* `sample` function

Recall that you can sample a bag of M&Ms using the following:

```{r}
## Colors as a vector
mm_colors <- c("blue", "brown", "green", "orange", "red", "yellow")

## Proportion/probability of each color
mm_probs <- c(.24, .14, .16, .20, .13, .14)

## I want to "sample" a bag of MMs
new_bag <- sample(x = mm_colors, size = 55, replace = TRUE, prob = mm_probs)
```

Also, here's how to read in our saved dataset, assuming it's in your `data` folder.

```{r}
mm_dataset <- read.csv("/Users/mlammens/Google Drive/Professional/Websites/Biostats/data/class_mm_data.csv", row.names = 1)
```

I can then get my bag of M&Ms using:

```{r}
my_bag <- mm_dataset["Matt",]
```


```{r}
## Set up a counter to count the number of times the simulated bag matches my bag,
## and start it at 0
count<-0

## Set the number of replications to try
replications<-3000

for(i in 1:replications){
  
  ## Draw a new bag
  new_bag <- sample(x = mm_colors, size = sum(my_bag), replace = TRUE, prob = mm_probs)

  ## Count the number of each color
  new_bag_counts <- c(sum(new_bag == "blue"),
                      sum(new_bag == "brown"),
                      sum(new_bag == "green"),
                      sum(new_bag == "orange"),
                      sum(new_bag == "red"),
                      sum(new_bag == "yellow"))

  ## Check to see if my bag is the same as the new bag
  if( all(my_bag == new_bag_counts) ){
    count<-count+1
    }
  }

  ## Calculate the probability of my bag based on the number of replications done
  probability<-count/replications

print(paste0("Likelihood of my bag: ", probability))
```

***

## Data exploration

The first step to any data analysis is to interrogate your data by calculating some standard statistics and by visualizing it it various ways. 
In future classes we will look more closely at standard statistics.
Today, we'll focus on data visualization.

### Iris dataset

We're going to work with a dataset that comes built in to R, commonly 
called the
[iris dataset](https://stat.ethz.ch/R-manual/R-patched/library/datasets/html/iris.html).
It is also sometimes called Fisher's Iris dataset (but should more appropriately
be called Anderson's Iris dataset). Because it comes pre-packaged with R, we 
can load it into our environment using the `data` function.

```{r}
data( iris )
```

Let's have a look at this dataset

```{r}
head( iris )
tail( iris )
```

The dataset contains measurements of four characteristics of three
different species of Iris (plants!). 

#### `summary` function

Let's begin by using the `summary` function to examine this dataset.
`summary` returns many of the standard statistics.
When doing data exploration, a few things you want to look at are:

* How to the mean and median values within a variable compare?
* Do the min and max values suggest there are outliers?
* Which variables (i.e., columns) are quantitative (numeric) versus categorical (factors or characters)

```{r}
summary(iris)
```

#### Aside: Differences between characters and factors

Factors 'appear' to be similar to characters, but are in fact coded numerically in R. 
Think of **factors** like **categories**. 
Here's a quick example that demonstrates the difference in these two variable types that shows up when using `summary`.

```{r}
## Make a new iris dataset
iris_new <- iris

## Create a new column that treats species as a character, rather than a factor
iris_new$species_char <- as.character(iris_new$Species)

## Run summary command
summary(iris_new)
```


#### Aside: A (very) brief introduction to navigating a `data.frame`

We will be very brief here. 
I recommend checking out 
[this Data Carpentry lesson](http://www.datacarpentry.org/R-ecology/03-data-frames.html) for more information.

* Looking at specific `data.frame` elements. Use the *row* and *column* notation.

Here is the 5th row, 3rd column (Petal.Length).
**Note: We are using square brackets to index the `data.frame` and we *always* use row, column notation.**

```{r}
iris[5, 3]
```

* Looking at an entire column.

Here are two ways to get the `Petal.Length` column.

First, **note: we leave the row part blank, but still add the comma.**

```{r}
iris[ ,3]
```

Second, **use only the variable (column) name. Note the use of the `$` operator**

```{r}
iris$Petal.Length
```

* Looking at specific column entry

This is another way to look at the 5th entry in the `Petal.Length` column.

```{r}
iris$Petal.Length[5]
```

* Looking at all entries for a given row.

Here's all the entries for the 5th row.
**Note: here we leave the column part blank, but still add the comma.**

```{r}
iris[5, ]
```

* Looking at a set of rows and/or columns.

Here's all the entries in the 5th through 10th rows, 1st through 3rd columns.
**Note: we use the `:` operator to look at a range of value.**

```{r}
iris[5:10, 1:3]
```

* For `data.frame`s, if you do not use row, column notation, you will get only the columns back.

```{r}
head(iris[2:3])
```

***

#### Challange

What am I going to get if I execute the command below?

```{r, results='hide'}
head(iris[c("Sepal.Width","Petal.Length")])
```

***

### Visualization using `base` R functions

Let's add a habitat type variable to the iris dataset. We'll use this later. 
*Caveat - I made up these habitat type preferences.*

```{r}
iris_habitat <- data.frame( Species = c( "setosa", "versicolor", "virginica" ),
                            Habitat = c( "forest", "wetland", "meadow" ) )

iris_full <- merge( x = iris, y = iris_habitat, by = "Species" )

head( iris_full )
tail( iris_full )
```

#### Visualizing the measurements of a single variable

Perhaps the most common way to look at data for a single variable is a histogram.
This essentially is a bar plot, where each bar represents the number of times a value falls within a particular *bin*.

Example: let's look at the distribution of `Petal.Length` values

```{r}
hist(iris_full$Petal.Length)
```

***

#### Challange - use `?hist` to determine how to change the number of bins used

Hint: bins are sometimes called breaks

```{r}
?hist
hist(iris_full$Petal.Length, breaks = 30)
```

***

#### Visualizing relationships between two variables

We would like to plot Sepal.Length versus Petal.Length. We'll first do this using
**base R** functions.

```{r plot1}
plot( x = iris_full$Sepal.Length, y = iris_full$Petal.Length )
```

Note, we could save a little typing by using the `with` function. This 
sets up a temporary environment where all of the variables (columns) 
in my dataset are defined individually.

```{r plot1-with}
with( data = iris_full, plot( x = Sepal.Length, y = Petal.Length ) )
```


### Visualization using the `ggplot2` package

Now we're going to introduce a data visualization package called **ggplot2**.
This package is great for producing publication quality graphics, but the syntax
used to create a plot is a little more involved.

#### Aside: Installing and loading packages

First, let's install the `ggplot2` package:

```{r installggplot2, eval=FALSE}
# Only need to do this once
install.packages("ggplot2")
```

Then load it:

```{r loadggplot2, message=FALSE}
library(ggplot2)
```


OK, let's fire up ggplot.

```{r plot1-ggplot}
ggplot() + # First make an empty ggplot canvas. Notice the trailing plus sign
  geom_point( data = iris_full, aes( x = Sepal.Length, y = Petal.Length ) )
```

Let's break down that call to introduce a few key things about ggplot

* ggplot: the initial canvas we're working on
* geom: geometric objects (i.e. the type of plot - histogram, points, line, etc) 
* aes: aesthetic mapping 

**THAT SEEMS SO COMPLICATED!**

It's true, for simple plots, ggplot can be much more complicated
than simply using the base functions. But the power of ggplot lies
in the ability to lay several geometries (geoms) over each other.
Also, each geometry has a rich set of options. For example,
let's say I want to create the plot we just made, but have 
each species represented by a different color.

```{r plot2}
ggplot() + # First make an empty ggplot canvas. Notice the trailing plus sign
  geom_point( data = iris_full, aes( x = Sepal.Length, y = Petal.Length, colour = Species ) )
```

Let's add more information - how about habitat type as well.

```{r}
ggplot() + # First make an empty ggplot canvas. Notice the trailing plus sign
  geom_point( data = iris_full, aes( x = Sepal.Length, y = Petal.Length, colour = Species, shape = Habitat), size = 2.5 ) +
  theme_bw()
```


**facets** - a way to separate data into different subplots

Let's say we wanted different plots for each species. We can do that in ggplot using
`facets`.

```{r plot2-facets}
ggplot( data = iris_full, aes( x = Sepal.Length, y = Petal.Length ) ) + 
  geom_point() +
  facet_grid( Species ~ . )
```

**NOTE**: I moved the `data ...` stuff into the initial `ggplot` call here.

*** 

#### Challange

1. `ggplot2` has many geometries, allowing us to make lot's of different 
types of plots. Let's make two new plots - one **boxplot** of `Petal.Length`, 
with one boxplot for each species. Use `geom_boxplot` for this.

```{r}
ggplot(data = iris_full, aes(x = Species, y = Petal.Length)) +
  geom_boxplot(aes(fill = Species)) +
  theme_bw()  
```

2. Make a histogram of Petal.width. 

```{r}
ggplot() +
  geom_histogram(data = iris_full, aes(x = Petal.Length)) +
  theme_bw()  
```


Use facets to separate the three species.

```{r}
ggplot() +
  geom_histogram(data = iris_full, aes(x = Petal.Length, fill = Species)) +
  facet_grid( Species ~ . ) +
  theme_bw()  
```

3. Second, let's make density plots of Petal.Width. Use `geom_density` and
`colour` to make different colored density lines for iris in each 
habitat type. 
**Note: the area under the curve is equal to 1.**

```{r}
ggplot() +
  geom_density(data = iris_full, aes(x = Petal.Length, colour = Species)) +
  theme_bw()
```

4. Use histogram to plot density instead of counts.

```{r}
ggplot() +
  geom_histogram(data = iris_full, 
                 aes(x = Petal.Length, y = ..density.., fill = Species)) +
  facet_grid( Species ~ . ) +
  geom_density(data = iris_full, aes(x = Petal.Length, colour = Species)) +
  theme_bw()  
```


***

### More `ggplot2` resources

* <http://docs.ggplot2.org/current/>: The official **ggplot2** documentation.
* <http://www.amazon.com/o/ASIN/0387981403/ref=nosim/gettgenedone-20>: The **ggplot2** book, by the developer, Hadley Wickham.
* <https://groups.google.com/forum/#!forum/ggplot2>: The **ggplot2** Google Group (mailing list, discussion forum).
* <https://github.com/swcarpentry/bc/tree/master/intermediate/r/data-visualization>: Intermediate Software Carpentry lesson on data visualization with **ggplot2**.
* <http://learnr.wordpress.com/>: A blog with a good number of posts describing how to reproduce various kind of plots using **ggplot2**.
* <http://stackoverflow.com/questions/tagged/ggplot2>: Thousands of questions and answers tagged with "ggplot2" on Stack Overflow, a programming Q&A site.

***


## Logistics

* **Office hours:** Two hours scheduled, Thursdays 4:00 to 6:00 PM. 
* **Additional hours:** Can meet by appointment on Mondays or Thursdays in person. By Skype/ Google Hangouts another day.
* **Homework:** Due by 5:39 PM the week *after* it was assigned, unless otherwise noted.

