---
title: "Sandbox"
output: html_document
---

# Probability

### Frequentist perspective 

The relative frequency of an event over the long term, after many trials.

E.g., throwing a coin. What's the probability of a 'heads'? After one trial, two trials, ..., 1000 trials.

```{r}
(head <- runif(1) > 0.5)
```


### Bayes perspective

Stay tuned.

***


## M&Ms

Open your bag and count how many M&Ms you got. Also count how many of each color.

### Assignment operator

Make a new variable (object) called `tot_mm` and assign it the number of MMs in your bag.

```{r}
tot_mm <- 60
```

### `mean` number of MMs in a bag

* Vector
* `sum` function
* `mean` function

Calculate the mean number of MMs in a bag.

```{r}
tot_mm_pop_sample <- c(60, 58, 57, 63, 61, 60, 59)
```

Mean:

$$
\bar{x} = \frac{\sum_{i=1}^n{x_i}}{n}
$$

Using the `sum` function.

```{r}
(mean_mm_1 <- sum(tot_mm_pop_sample) / length(tot_mm_pop_sample))
```

Using the `mean` function.

```{r}
mean_mm_2 <- mean(tot_mm_pop_sample)
```

### Quantify the variation in the number of MMs in a bag

* `var` function
* `sd` function

What do we mean by variation? 
In general, we mean something along the lines of 'the amount of variation around some mean value'.

**How might we quantify this?**

* Add up the distances from the mean. 

First calculate distances from the mean for each bag.

```{r}
(dist_from_mean <- tot_mm_pop_sample - mean_mm_1)
```

Next calculate the total distances.

```{r}
(sum_dist_from_mean <- sum(dist_from_mean))
```

What information does this give us, and why might it not be useful?

* Add up the *absolute* distances form the mean. 

First calculate the *absolute* distances from the mean.

```{r}
(absdist_from_mean <- abs(tot_mm_pop_sample - mean_mm_1))
```

Next add up the total *absolute* distances from the mean.

```{r}
(sum_absdist_from_mean <- sum(absdist_from_mean))
```

* Add up the *squared* distances from the mean.

While the absolute distance from the mean does give us a reasonable measure of variability, because of specific mathematical properties, it's more convenient to work with *squared* distances from the mean, leading to the measures of **variance** and **standard deviation**.

**Challange**

Calculate *squared* distances from the mean, and sum them to determine the total squared deviation.

**Challange**

Can you find an R function to do this in one step?

**Standard deviation** can be thought of as the average deviation from the mean.

## Probability and M&Ms

Without looking, you chose one M&M. What colors could you have chosen?

The **set** of brown, yellow, green, red, orange, blue is the **sample space**.

Your selection of a single M&M is called an **event**.


**Challange**

* What's the probability of getting a "color" M&M in **your** bag?

P(brown), P(yellow), etc.

**Challange**

What is the mean probability of getting each color across each bag?

Tools to use:

* `data.frames`
* `apply`


## Venn diagrams

* What's the P(green OR blue OR red) in **your** bag? 

P(green) + P(blue) + P(red)

* What is the P(NOT green) in **your** bag?

$$
P(NOT Green) = 1 - P(Green) = P(G)^c
$$


### Sampling with replacement

* What's the P(green and then blue), first putting back your first draw?

Intersection: P(green intersection blue) = P(green) * P(blue) 

* What's the P(green or blue)?

Union: P(green) + P(blue) - P(green and blue)


### Sampling without replacement

* What is the probability of getting green, then blue, *without replacing your first draw*?

**Challange**

What is the sample space for drawing two M&Ms?

Well work further on similar problems in future classes.


## Draw a random bag of M&Ms using the company stated frequencies/probabilities for each color

```{r}
## Colors as a vector
mm_colors <- c("blue", "brown", "green", "orange", "red", "yellow")
mm_probs <- c(.24, .14, .16, .20, .13, .14)

## I want to "sample" a bag of MMs
new_bag <- sample(x = mm_colors, size = 55, replace = TRUE, prob = mm_probs)
table(new_bag)

```

