Problem Set 4: Linear Regression
========================================================

**Suggested submission: Apr 4, 2016**

For this problem set we will use data on cell survival for fourteen plates exposed to different levels of radiation.
You can acquire these data by using the following command:

```{r}
cell_survival <- read.csv(file = "https://mlammens.github.io/Biostats/data/cell_survival.csv")
```

## Part 1

1. Use the `lm` function to fit a linear regression model to these data. 
Report the results and plot the data and the best-fit line. (5 pnts)

2. Are there any data points that may be considered outliers? 
Use the regression diagnostics we learned about in class to examine these data.
If there are outliers, how does the regression slope change if it/they were to be removed from the dataset? 
Be sure to clearly state which data points you are considering "outliers" and why. (5 pnts)

3. Do these data fit the assumptions of linear regression?
Consider each of the "big 3" assumptions separately. (5 pnts)

## Part 2

Instead of using `lm`, we can write a function to calculate the negative log-likelihood associated with fitting a linear regression model. 
In order to do this, let's consider the linear regression model as we introduced it on page 1 of the written notes from Lecture 7:

$$
Y \sim a + bX + \epsilon
$$

where $\epsilon = N(0,\sigma^2)$, and this can be equivalently written as:

$$
Y \sim N( a + bX, \sigma^2)
$$

To write the negative log-likelihood function for our observed data, we can use the following R code:

```{r}
survival <- cell_survival$log_survival_prob
dose <- cell_survival$dose_rads_per_100
neg.ll <- function( params ){
  -sum( dnorm( survival, mean=(params[1]+params[2]*dose), sd = params[3], log=TRUE) )
}
```

1. Describe in words what this function is "doing". 
That is, in what ways is this function providing us with the negative log-likelihood of these data? 
(**Hint:** You might want to describe what each piece of the R code is for.) (5 pnts)

2. Using the `optim` function in R, minimize the negative log-likelihood to obtain the maximum likelihood regression parameter estimates for ($\beta_0$), ($\beta_1$), and ($\sigma^2$).
(**Hint 1:** Have a look back at [Lecture Notes - 5](http://mlammens.github.io/Biostats/lectures/Lecture-5.html) for an example of using `optim`.
**Hint 2:** You will need starting values for optim; the results of the `lm` calculation would be reasonable starting points.) (5 pts)
