---
title: "Problem Set 1"
author: "Matthew E. Aiello-Lammens"
date: '2018-01-24'
output: html_document
---

**Please submit this homework as an R Markdown (Rmd) file.**

As I mentioned during our first meeting, rmarkdown is a way to incorporate text, with simple formatting, with R code chunks.
Rather than spell out how to create an Rmd file here, I recommend you consult the following resources:

* [RStudio Rmarkdown Quick Tour](http://rmarkdown.rstudio.com/authoring_quick_tour.html)
* [RStudio Rmarkdown Intro Article](http://rmarkdown.rstudio.com/articles_intro.html)
* [OurCodingClub Intro Example](https://ourcodingclub.github.io/2016/11/24/rmarkdown-1.html)

#### Your file should use the following naming scheme

**`[last name]_ENS623_SP18_PS1.Rmd`**

For example, 

`Lammens_ENS623_SP18_PS1.Rmd`

#### Grading note

Problems 1-3 are each worth 5 points, Problem 4 is worth 10.

## Problem 1

Using the help features in R, i.e. preface a function with `?`, look up how to use the `matrix` function. 
Create two different matrices that match those presented below.

#### Matrix 1

```{r, echo=FALSE}
(matrix_1 <- matrix(data = 1:9, nrow = 3, byrow = TRUE))
```

#### Matrix 2

```{r, echo=FALSE}
matrix(data = 1:9, nrow = 3, byrow = FALSE)
```


## Problem 2

In R, you can access individual elements of a vector, matrix, data frame, etc. using the square brackets, `[]`. 
In the example below, I create a vector of pets, and access individual and sets of elements.


```{r}
# Create my pets vector
pets <- c("cat", "dog", "rabbit", "pig", "snake")

# Access the first elemetn
pets[1]

# Access multiple elements, in sequence
pets[3:4]

# Access multiple elements, out of sequence
pets[c(1,4)]
```

**Use this knowledge regarding accessing elements to reverse the order of the `pets` vector.**


## Problem 3

You can access elements of a matrix using **row**, **column** notation.
For example, below I access the middle element of **Matrix 1** above.

```{r}
matrix_1[2,2]
```

**Use this to make a new 2-row by 2-column matrix comprised of elements from Matrix 1**


#### Bonus

Re-assign the center value of the matrix to a value of 99.

## Problem 4

Read Chapter 1 of *How to do Ecology* and skim the Sutherland et al. 2013 paper (both posted on BlackBoard). 
In less than 250 words, respond to these readings. 
Some prompts that might help you in your response (though you are not obliged to use these) -

* Considering your short- and medium-term goals, are you concnered with asking a "novel" question?
* What is the relationship between ecological questions and those that are important in environmental science?
* Could you sketch out a question you are interested in tackling?
* Are there particular topics, or questions, in the Sutherland paper you find inspiring?
