---
title: "Lecture 7 - Linear Models, Correlation, and Regression"
author: "Matthew E. Aiello-Lammens"
output: 
  html_document: 
    toc: yes
---

# Logistics

* Solution sets for problem sets 1-3 will be posted on the Open Science Framework site on 3/21. This will also be the last day to submit any past assignments.
* The take-home midterm will be distributed tomorrow afternoon.
* The midterm is due Monday, 3/28.

# Relationships among the distributions

![distribution relationships](/Volumes/Garage/Professional-Annex/SBU-Grad-School/SBU-Teaching/Biometry-BEE-552-SP13/Figure adapted from Leemis (1986).pdf) 

# Linear model review

**Written Notes - Pages 1-3**

## Co-variation

Calcualte covariation between sepal length and sepal width, using the `iris` dataset.

```{r}
data(iris)
head(iris)

library(ggplot2)
qplot(x = iris$Sepal.Length, y = iris$Sepal.Width, colour = iris$Species) +
  theme_bw()

covar <- sum((iris$Sepal.Length - mean(iris$Sepal.Length)) * (iris$Sepal.Width - mean(iris$Sepal.Width)))/ (nrow(iris) - 1)

## Confirm
var(x = iris$Sepal.Length, y = iris$Sepal.Width)

```


**Written Notes - Page 4**

## Correlation

Calculate the correlation between sepal length and sepal width, and calculate whether $r$ is significantly different from 0 or not.

```{r}
cor_iris <- covar / (sd(iris$Sepal.Length)*sd(iris$Sepal.Width))

se_cor_iris <- sqrt((1 - cor_iris^2)/(nrow(iris) - 2))

t_cor_iris <- cor_iris/se_cor_iris

## Calculate the probability of getting our t value, or one more extreme
pt(q = t_cor_iris, df = (nrow(iris)-2))*2

## Use `cor` to find correlation value
with(data = iris, cor(x = Sepal.Length, y = Sepal.Width, method = "pearson"))

## Test the correlation
with(data = iris, cor.test(x = Sepal.Length, y = Sepal.Width, method = "pearson"))


```


## Robust correlation

For Pearson's correlation coefficient to be appropriate, the data should have:

1. A linear relationship
2. A bivariate normal distribution

* Spearman's Rank Correlation - calculates correlation on ranks of observed values
* Kendall's Rank Correlation


# Linear regression

Now we assume that any change in **y** is due to a change in **x**.

## Example of linear regression

Effects of starvation and humidity on water loss in the confused flour beetle.
Here, looking at the relationship between humidity and weight loss

```{r}
flr_beetle <- read.csv(file = "https://mlammens.github.io/Biostats/data/Logan_Examples/Chapter8/Data/nelson.csv")
flr_beetle
```

Plot these data

```{r}
ggplot() +
  geom_point(data = flr_beetle, aes( x = HUMIDITY, y = WEIGHTLOSS)) +
  theme_bw()
```

Run a linear regression

```{r}
flr_beetle_lm <- lm(data = flr_beetle, WEIGHTLOSS ~ HUMIDITY)

## This will give us a multitude of diagnostic plots
plot(flr_beetle_lm)

summary(flr_beetle_lm)
```

Plot these data, with `lm` fit

```{r}
ggplot(data = flr_beetle, aes( x = HUMIDITY, y = WEIGHTLOSS)) +
  geom_point() +
  stat_smooth(method = "lm") +
  theme_bw()
```

