---
title: "Lecture 10 - Multiple And Complex Regression"
author: "Matthew E. Aiello-Lammens"
output: 
  html_document: 
    toc: yes
---

# Challange

Find the best fit model for the data on the spatial spread of Glossy Buckthorn.

```{r}
buckthorn <- read.csv(file = "https://mlammens.github.io/Biostats/data/buckthorn.csv")
```


# Non-linear regression

Species-area relationship curves are often observed to have some sort of **Power Law** relationship:

*Number of species = a constant \* Area^z^*

Peake and Quinn (1993) examined this relationship for invertebrates living in inter-tidal mussel clumps (Logan - Example 9G; Quinn and Keough - Box 6.11).

Get the data.

```{r}
sar <- read.csv(file = "https://mlammens.github.io/Biostats/data/Logan_Examples/Chapter9/Data/peake.csv")
head(sar)
summary(sar)
```

Visualize these data.

```{r}
library(ggplot2)

ggplot(data = sar, aes(x = AREA, y = SPECIES)) +
  geom_point() +
  theme_bw() +
  stat_smooth()
```

Naively fit a linear model to these data.

```{r}
sar_lm <- lm(data = sar, SPECIES ~ AREA)
plot(sar_lm)
```

Now let's use `nls` to fit a power law model.

```{r}
sar_nls <- nls(data = sar, formula = SPECIES ~ const * AREA^z, 
               start = list(const = 0.1, z = 1))
summary(sar_nls)
```

Make a residuals versus fitted plot for these data

```{r}
plot(resid(sar_nls) ~ fitted(sar_nls))
```

Compare the linear and non-linear models using AIC.

```{r}
AIC(sar_lm)
AIC(sar_nls)
```

Briefly review Table 9.1 on page 212 of Logan.


# ANOVA

## Review from [Lecture 9](http://mlammens.github.io/Biostats/lectures/Lecture-9.html)


## Example

From Medley and Clements (1998), by way of Logan 2010 - investigation of the impact of zinc contamination on diversity of diatom species in the USA Rocky Mountains.

```{r}
diatoms <- read.csv(file = "https://mlammens.github.io/Biostats/data/Logan_Examples/Chapter10/Data/medley.csv")
```


```{r}
head(diatoms)

summary(aov(data = diatoms, DIVERSITY ~ ZINC ))
anova(lm(data = diatoms, DIVERSITY ~ ZINC ))


```